\vspace{-2cm} Inference on the Effects of Observed Features \\ in Latent Space Models for Networks

### 2017-05-07 manuscript.tex
making final edits

\noindent Due to the complex interdependence found in relational data, political networks scholars draw upon an increasingly sophisticated toolkit for statistical inference. The latent space model (LSM) for network data combines a generalized linear model with a latent spatial embedding of the network. It is assumed that the latent spatial embedding can control for unmeasured confounding structure that is related to the values of edges in the network. There has been little to no research that considers the LSM's performance in adjusting for unmeasured structure. We investigate the LSM's performance via a simulation study. In the presence of an unmeasured covariate that can be modeled using a latent space, estimation and inferential error remain high under even moderate confounding. However, the prediction error of the LSM when unmeasured network structure is present is substantially lower in most cases. We conclude that the LSM is most appropriate for exploration or prediction.\footnote{This work was supported by National Science Foundation grants DGE-1144860, SES-1558661, SES-1619644, and CISE-1320219. Any opinions, findings, and conclusions or recommendations are those of the authors and do not necessarily reflect those of the sponsor.}

@@

Recent work on the LSM has explicitly addressed the potential problem of Type 2 error. The AMEN framework for latent variable modeling of networks is explicitly designed to model structure in the residuals---structure that is leftover after accounting for the observed covariates. \citet[p. 43]{hoff2015dyadic} notes that the latent factors represent patterns in the network that, ``aren't explained by the known regressors.'' \citet[pp. 12--13]{minhas2016inferential} also describe how AMEN is designed such that the multiplicative effects, ``capture higher-order dependence patterns that are left over [in the stochastic linear predictor] after accounting for any known covariate information.''  The approach incorporated in AMEN is effective at avoiding correlation between the observed covariates and latent factors, which would result in Type 2 error. However, there have been no methodological innovations to avoid Type 1 error in latent variable modeling with networks. Below we review several studies in which researchers have relied on the latent space modeling framework to adjust for unmeasured confounders, and through a simulation study we show that the LSM as implemented in the R package \texttt{latentnet} as well as \texttt{amen} performs poorly in terms of Type 1 error when the unmeasured network structure confounds the relationship between the observed covariate and the network.

@@

We also repeat this simulation study using \texttt{amen} \citep{hoff2015dyadic, minhas2016inferential}. We omit the poisson family, and, due to the computational cost of the simulation, we consider only networks with 25 nodes or 100 nodes. Rather than constructing an omitted variable as described above, we draw from a standard normal distribution $\mathbf{u}$, and, because we are considering symmetric networks, take the outer product of $\mathbf{u}$ with itself to construct the latent factor. We take 100,000 MCMC samples after 500 burn-in iterations, and we assess convergence using the diagnostic of \cite{raftery1992practical}.

@@

To evaluate estimation error we compute the bias. Figures \ref{fig:estimation_ls} and \ref{fig:estimation_nls} shows the results. In Figure \ref{fig:esimation_ls} we can see that with a uniform distribution over the correlation between the distances and the observed covariate, all methods (except the true model wherein the distances are included via an observed covariate) show substantial bias. In most cases the LSM performs better than the GLM, sometimes by as much as $50\%$. However, in absolute terms, the amount of bias is large. Our results demonstrate that using the LSM will not permit the discovery of and adjustment for the true latent positions when the observed covariates and latent distances are confounded.  For both the GLM and the LSM, bias decays rapidly as the degree of correlation between the distances and the measured covariate decreases. In \ref{fig:estimation_nls} we can see that when there is no omitted network structure, the LSM exhibits bias greater than that of the GLM.

@@

\begin{figure}
\includegraphics[width=\textwidth]{figures/generalization.png}
\caption{The $x$-axis shows a Monte Carlo estimate of the mean square error for edge values drawn from the appropriate distribution with the observed covariate $\mathbf{x}$ and the latent positions $\mathbf{z}$ fixed. In the Bernoulli case the Brier score is computed and in the poisson and normal cases the mean-square-error. \label{fig:generalization}}
\end{figure}

In our repetition of this simulation with \texttt{amen} we find similar results, with the exception of type-2 error rates, which, as expected, are substantially lower for \texttt{amen}. However, type-1 error rates remain elevated, as is bias in the case where there is a latent dependence structure that is correlated with the observed covariate. \texttt{amen} does perform better in the case where it is used despite the absence of such a latent dependence structure though, relative to \texttt{latentnet}, which is to be expected given that it outperforms \texttt{latentnet} in terms of type-2 error. Again, \texttt{amen} performs well in terms of generalization error, coming close to the theoretical lower bound. The results of our simulations with \texttt{amen} are available in the Online Appendix.

@@

\begin{figure}
\includegraphics[width=\textwidth]{figures/generalization.png}
\caption{The $x$-axis shows a Monte Carlo estimate of the mean square error for edge values drawn from the appropriate distribution with the observed covariate $\mathbf{x}$ and the latent positions $\mathbf{z}$ fixed. In the Bernoulli case the Brier score is computed and in the poisson and normal cases the mean-square-error. \label{fig:generalization}}
\end{figure}

>>>>

\noindent Due to the complex interdependence found in network data, political networks scholars draw upon an increasingly sophisticated toolkit for statistical inference. The latent space model (LSM) for network data combines a generalized linear model with a latent spatial embedding of the network. In numerous applications, researchers assume that the latent spatial embedding can control for unmeasured confounding structure that is related to the values of edges in the network. There has been little to no research that considers the LSM's performance in adjusting for unmeasured structure. We investigate the LSM's performance via a simulation study. In the presence of an unmeasured covariate that can be modeled using a latent space, estimation and inferential error remain high under even moderate confounding. However, the prediction error of the LSM when unmeasured network structure is present is substantially lower in most cases. We conclude that the LSM is most appropriate for exploration or prediction.%\footnote{This work was supported by National Science Foundation grants DGE-1144860, SES-1558661, SES-1619644, and CISE-1320219. Any opinions, findings, and conclusions or recommendations are those of the authors and do not necessarily reflect those of the sponsor.}

@@

Recent work on the LSM has explicitly addressed the potential problem of Type 2 error. The AMEN framework for latent variable modeling of networks is explicitly designed to model structure in the residuals---structure that is leftover after accounting for the observed covariates. \citet[p. 43]{hoff2015dyadic} notes that the latent factors represent patterns in the network that, ``aren't explained by the known regressors.'' \citet[pp. 12--13]{minhas2016inferential} also describe how AMEN is designed such that the multiplicative effects, ``capture higher-order dependence patterns that are left over [in the stochastic linear predictor] after accounting for any known covariate information.''  The approach incorporated in AMEN is effective at avoiding correlation between the observed covariates and latent factors, which would result in Type 2 error. However, there have been no methodological innovations to avoid Type 1 error in latent variable modeling with networks. Below we review several studies in which researchers have relied on the latent space modeling framework to adjust for unmeasured confounding structure.  Through a simulation study we find that the LSM, as implemented in both the R \citep{R} packages \texttt{latentnet} \citep{latentnet} and \texttt{amen} \citep{amen}, performs poorly in terms of Type 1 error when the unmeasured network structure confounds the relationship between the observed covariate and the network.

@@

We also repeat this simulation study using \texttt{amen} \citep{hoff2015dyadic, minhas2016inferential}. We omit the Poisson family, and, due to the computational cost of the simulation, we consider only networks with 25 nodes or 100 nodes. Rather than constructing an omitted variable as described above, we draw from a standard normal distribution $\mathbf{u}$, and, because we are considering symmetric networks, take the outer product of $\mathbf{u}$ with itself to construct the latent factor. We take 100,000 MCMC samples after 500 burn-in iterations, and we assess convergence using the diagnostic of \cite{raftery1992practical}.

@@

To evaluate estimation error we compute the bias. Figures \ref{fig:estimation_ls} and \ref{fig:estimation_nls} shows the results. In Figure \ref{fig:estimation_ls} we can see that with a uniform distribution over the correlation between the distances and the observed covariate, all methods (except the true model wherein the distances are included via an observed covariate) show substantial bias. In most cases the LSM performs better than the GLM, sometimes by as much as $50\%$. However, in absolute terms, the amount of bias is large. Our results demonstrate that using the LSM will not permit the discovery of and adjustment for the true latent positions when the observed covariates and latent distances are confounded.  For both the GLM and the LSM, bias decays rapidly as the degree of correlation between the distances and the measured covariate decreases. In \ref{fig:estimation_nls} we can see that when there is no omitted network structure, the LSM exhibits bias greater than that of the GLM.

@@

\begin{figure}
\includegraphics[width=\textwidth]{figures/generalization.png}
\caption{The $x$-axis shows a Monte Carlo estimate of the mean square error for edge values drawn from the appropriate distribution with the observed covariate $\mathbf{x}$ and the latent positions $\mathbf{z}$ fixed. In the Bernoulli case the Brier score is computed and in the Poisson and normal cases the mean-square-error. \label{fig:generalization}}
\end{figure}

In our repetition of this simulation with \texttt{amen} we find similar results, with the exception of type-2 error rates, which, as expected, are substantially lower for \texttt{amen}. However, type-1 error rates remain elevated, as is bias in the case where there is a latent dependence structure that is correlated with the observed covariate. \texttt{amen} does perform better in the case where it is used despite the absence of such a latent dependence structure though, relative to \texttt{latentnet}, which is to be expected given that it outperforms \texttt{latentnet} in terms of type-2 error. Like the LSM estimated with \texttt{latentnet}, \texttt{amen} performs well in terms of generalization error. The results of our simulations with \texttt{amen} are available in the Online Appendix.

@@

\begin{figure}
\includegraphics[width=\textwidth]{figures/estimation_ls_amen.png}
\caption{The bias of estimates of the effect of the observed covariate $\mathbf{x}$ when there is an ommitted variable. The $x$-axis gives the value of the parameter $\eta$ which controls the degree of dependence between $\mathbf{x}$ and ommitted covariate. Lower values of $\eta$ indicate higher levels of dependence between the observed and ommitted covariate. The $y$-axis gives a Monte Carlo estimate of the bias. The number of nodes are indicated in the top panels, while the distributional family of the edges is shown on the right panel. Each panel represents 4 values of $\eta$ with 1,000 Monte Carlo iterations executed at each point.
\label{fig:estimation_ls_amen}}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{figures/estimation_nls_amen.png}
\caption{The bias of estimates of the effect of the observed covariate $\mathbf{x}$ when there is no ommitted variable.
\label{fig:estimation_nls_amen}}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{figures/inference_type_1_amen.png}
\caption{Monte Carlo estimates of the Type-1 error regarding the effect of the observed covariate $\mathbf{x}$ are shown on the $y$-axis. Here, $\beta = 0$ and the error rate shown in each panel in that row gives 1 minus the probability of a 95\% confidence region (for the LSM) or interval (for the GLM) including $0$, giving the probability that a true null hypothesis of $\beta = 0$ is falsely rejected. \label{fig:inference_type_1_amen}}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{figures/inference_type_2_amen.png}
\caption{Monte Carlo estimates of the Type-2 error regarding the effect of the observed covariate $\mathbf{x}$ are shown on the $y-axis$. Here, $\beta = 1$, and the error rate shown in each panel in each row gives the probability of the probability/confidence intervals covering $0$, giving the probability of accepting a false null hypothesis. \label{fig:inference_type_2_amen}}
\end{figure}

\begin{figure}
\includegraphics[width=\textwidth]{figures/generalization_amen.png}
\caption{The $x$-axis shows a Monte Carlo estimate of the mean square error for edge values drawn from the appropriate distribution with the observed covariate $\mathbf{x}$ and the latent positions $\mathbf{z}$ fixed. In the Bernoulli case the Brier score is computed and in the normal case the mean-square-error. \label{fig:generalization_amen}}
\end{figure}
